Program Optimisations via Hylomorphisms for Extraction of Executable Code

### 2025-06-27 paper/paper.tex
Types

  This paper presents the first Coq formalisation of \emph{hylomorphisms}
  allowing for the mechanisation of all recognised recursive algorithms. The key
  contribution of this paper is that this formalisation is fully
  \emph{axiom-free}, and it allows the extraction of safe, idiomatic functional
  code. We exemplify the framework by formalising a series of algorithms based
  on different recursive paradigms such as divide-and conquer, dynamic
  programming, and mutual recursion and demonstrate that the extracted
  functional code for the programs formalised in our framework is efficient,
  humanly readable, and runnable. Furthermore, we show the use of the
  machine-checked proofs of the laws of hylomorphisms to do program
  optimisations.
\end{abstract}

@@

To preserve the generality of hylomorphisms, we encode data structures using
polynomial functors. To ensure that the extracted code is runnable, avoid using
indexed types in the definition of our recursion schemes. To strengthen the
results of this paper, we avoid all axioms.

We list the contributions of this paper. First, in
Section~\ref{sec:category-theory}, we provide the theoretical background on
recursion schemes and hylomorphisms. We then provide a Coq formalisation of
\emph{hylomorphisms} (Section~\ref{sec:recursion-schemes}) that (1) is
\emph{fully axiom-free}; (2) allows the extraction of idiomatic and efficient
code; and (3) can use regular Coq equalities to do program calculation, derive
correct implementations, and apply optimisations. In
Section~\ref{sec:extraction}, we apply this framework to formalise practical
examples of divide-and-conquer, dynamic programming, and mutual recursion
algorithms. Additionally, we verify the short-cut fusion optimisation and
present the extracted optimised code.

@@

Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow $C \to
A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if it
satisfies
\begin{equation}
  \label{eq:hylo}
  h = a \circ F(h) \circ c
\end{equation}
A solution to this equation does not exist for an arbitrary algebra and
coalgebra pair and, in fact, a definition like this cannot be accepted by Coq.

@@

An example of a recursive coalgebra is the \coq{partition} function used in
quicksort, which has the type $\text{List } A \to B(\text{List } A)$, where the
functor $B$ is defined as $BX = X \times A \times X$. This function deconstructs
a list by selecting a pivot element of type $A$ and splitting the remaining
elements into two sublists.

Notice that the initial algebra for lists corresponds to the functor
$F X = 1 + A \times X$ and its associated algebra map (or its inverse, viewed as
a coalgebra), $\InOp$ has the type $\text{List } A \to F(\text{List
} A)$. Importantly, \coq{partition} is not this map, which means that it does
not arise from the standard initial algebra structure, and thus catamorphisms
here cannot be used to define recursive functions.

@@

For example, in the category \textbf{Set}, consider the polynomial functor
corresponding to
\[
F(X) = A + B \times X + C \times X^2.
\]
We can model this functor by using a \emph{container}~\cite{AbbottAG05}. This is a set $S$
of \emph{shapes}, for example, $S = A + B + C$ for the example above, together with a function $P(s)$ denoting the exponent of $X$ for each shape.
Then a container extension is defined as the sum of all maps
$P(s) \to X$
\[
F(X) = \sum_{s \in S} (P(s) \to X)
\]
The functorial action
of $F$ is given by post-composition, mapping an element \( (s, g) \) in
\( F(X) \) to \( (s, f \vcomp g ) \) in \( F(Y) \), for $f : A \to B$.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
%Recursion schemes provide an abstract way to consume and generate data.
We now consider the formalisation of hylomorphisms in Coq. We first formalise
container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}). In
Section~\ref{sec:coq-coalg} we formalise coalgebras and put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}
A direct encoding of containers as presented in
Section~\ref{sec:category-theory} requires the use of the functional
extensionality axiom, and axiom K to deal with equalities of objects of type
$P(s) \to X$. Axiom K states that a predicate on equality proofs holds for all
such proofs if it holds for reflexivity. This would be needed to reason about
the equality of any two $f : P(s_1) \to X$ and $g : P(s_2) \to X$, when $s_1 =
s_2$. In Coq, if we know $(s_1 , f) = (s_2 , g)$, we cannot extract a proof that
$f = g$ unless we assume axiom K, or the equality on the type of $s_1$ and $s_2$
is decidable.  Furthermore, upon extraction Coq will need to insert unsafe casts
for
anything of type $P(s)$, because OCaml requires the input value to have a type that is identical to the function parameter, and
in the general case this cannot be done in OCaml.
To avoid these problems, we use \emph{setoids}, and
encode families of positions using \emph{decidable validity predicates}.

@@

The \coq{LFix F} setoid equality is the least fixed point of the \coq{App F}
setoid equality, i.e. if \coq{LFixR F} represents the setoid equality of
\coq{LFix F}, and we have two \coq{x y : LFix F}, then \coq{LFixR F x y} iff
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p'
   -> LFixR F (cont x p) (cont y p'))
\end{coqcode}
We define smart constructors for the isomorphism of least fixed
points as proper morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}
Catamorphisms are constructed so that they structurally deconstruct the
datatype, call themselves recursively, and then compose the result using an
\coq{F}-algebra.
\begin{coqcode}
Definition cata_f (a : Alg F A) : LFix F -> A
  := fix f (x : LFix) := match x with
        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
        end.
\end{coqcode}
It is easy to prove that catamorphisms respect setoid equalities.  In fact, we
define it as a setoid morphism between the setoid of \coq{F}-algebras and the
setoid of functions from \coq{LFix F} to \coq{A}:
\begin{coqcode}
cata : forall `{setoid A}, Alg F A ~> (LFix F ~> A)
\end{coqcode}
Finally, we prove that catamorphisms satisfy their universal property (see
Section~\ref{sec:category-theory}):
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

@@

As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra $a$ and
$F$-coalgebra $c$, $f$ is a hylomorphism if it satisfies
\[
  f = a \circ F\; f \circ c
\]
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, a recursive function definition
like this cannot be directly accepted by Coq.

@@

% Note that such proofs will be erased when extracting code to OCaml, and this is
% standard practice when building general recursive functions in
% Coq\footnote{\dcas{This technique had a name in the community, but I cannot
% remember. Something predicates}}. The core definitions are the following:
% \begin{coqcode}
% Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

% \end{coqcode}
% Here, \coq{RecF_inv} provides the structurally smaller element for
% the Coq recursive definition. We then couple coalgebras with a proof that they
% are recursive, and add a coercion from recursive coalgebras to coalgebras.
% \begin{coqcode}
% Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
% \end{coqcode}
% In some cases, proving that coalgebras can only be applied finitely many
% times is straightforward (e.g.\ they are recursive on some inductive type in
% Coq). However, in many cases, proving that a coalgebra is recursive directly
% can be tedious and cumbersome. To ease development, we prove that if a
% given coalgebra respects a \emph{well-founded relation}, then it is
% necessarily recursive.
% \begin{coqcode}
% Definition respects_relation `{setoid A} (c : Coalg F A)
%       {B} (m : A -> B) (R : B -> B -> Prop)
%   := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).

@@

\section{Extraction}
\label{sec:extraction}
In this section we show how programs and their optimisations can be encoded in
our framework and how can they be extracted to idiomatic functional code.
Although our examples use OCaml as target language, extraction works equally
well to other target languages (e.g. Haskell or Scheme). In particular, we show
how to formalise sorting algorithms by looking at quicksort
(Section~\ref{sec:sorting}), we show how to formalise dynamic programming
algorithms by looking at knapsack (Section~\ref{sec:knap}), and, finally, we
show an example of the shortcut deforestation optimisation
(Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

The proof is by induction on the fact that \coq{qsplit} is recursive. This means
that \coq{qsplit} terminates on \coq{l}.  In practice this is done by using the
tactic \coq{induction (recP qsplit l)}. Then we can unfold \coq{hylo merge qsplit} to
\coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
uniqueness (Section~\ref{sec:coq-rec-coalgebras}). The rest of the proof is standard.
Overall, our proof is shorter than alternative implementations, and of
comparable complexity%
%% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
\footnote{The QSort definition and correctness proof in the
equations
package is under 200LOC --
\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
and other online proofs of correctness are of similar complexity
\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
% and can be found in the
% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

@@

% Mergesort is a hylomorphism with the recursive structure defined by the
% container \coq{(TreeF (list int) unit)}.
% The computationally expensive part of the Mergesort implementation is the
% algebra, since this is where most of the comparisons happen. The recursive
% coalgebra for Mergesort is defined as the following Coq functions:
% \begin{coqcode}
% Fixpoint splitL (x : list int) (accL accR : list int) :=
%   match x with
%   | nil => (accL, accR)
%   | cons x xs => splitL xs accR (cons x accL)
%   end.

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism, and
\coq{times_two} multiplies every element of the list by two. We can use Coq's
generalised rewriting~\cite{Sozeau09} and \coq{hylo_fusion_l} to fuse \coq{times_two} into
\coq{hylo merge qsplit} which gives the program \coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by applying
\coq{times_two} to the shapes thus multiplying every pivot by two.

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
We now showcase the shortcut deforestation optimisation on lists.  This is used
to remove the creation of intermediate data structures by fusing multiple
recursive traversals into one.  For example, we can use it to show that the following
code from Takano and
Meijer~\cite{TakanoM95}, which uses three different hylomorphisms, can be fused into one:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here \coq{sf1} is the composition of \coq{length},
\coq{Lmap f} and \coq{append ys} which are, in particular,
catamorphisms.
To prove our goal we use the so called \emph{acid rain} theorem~\cite{TakanoM95}
which states that given a parametric function of the following type:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
we can conclude, by parametricity, that the following equation holds:
\begin{coqcode}
  hylo a l_out \o hylo (s l_in) c =e hylo (s a) c
\end{coqcode}
Unfortunately, this is not provable in Coq without adding the parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}, but we can prove it for specific
functors, e.g. the functor of lists.

We prove a specialised version of the acid rain theorem that fuses
hylomorphisms defined in terms of the following function \coq{tau} that maps
list algebras to list algebras.  This essentially is a function on lists which
uses the algebra for both the recursive step and for the base case where it is
used in the hylomorphism to continue the recursion:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
\end{coqcode}
Our acid rain theorem is stated as follows:
\begin{coqcode}
  hylo a l_out \o hylo (tau l l_in) c =e hylo (tau l a) c
\end{coqcode}

Now we use \coq{tau} to define the \coq{append} function as follows:
\begin{coqcode}
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. The acid rain theorem allows us to fuse this definition
of append with the maps \coq{Lmap} and \coq{length} which is the optimisation we
wanted in the first place. The following is the extracted OCaml code from the
optimised program:
\begin{ocamlcode}
let rec sf1 f ys =
  function | [] -> let rec f0 = function
                   | [] -> (Uint63.of_int (0))
                   | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                   in f0 ys
           | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}
Notice how the \coq{Lmap} is optimised away and the function \coq{length} is
applied to both the arguments of the append function.

@@

Our final container extension is the following:
\coq{Sum (Const nat) (Prod Id Id)}.
The shape of this functor is:
\coq{nat + unit * unit}. The positions of this functor
are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
to \coq{unit + unit}, which will only be valid in the RHS.

>>>>

  This paper presents the first Rocq formalisation of \emph{hylomorphisms}
  allowing for the mechanisation of all recognised recursive algorithms. The key
  contribution of this paper is that this formalisation is fully
  \emph{axiom-free}, and it allows the extraction of safe, idiomatic functional
  code. We exemplify the framework by formalising a series of algorithms based
  on different recursive paradigms such as divide-and conquer, dynamic
  programming, and mutual recursion and demonstrate that the extracted
  functional code for the programs formalised in our framework is efficient,
  humanly readable, and runnable. Furthermore, we show the use of the
  machine-checked proofs of the laws of hylomorphisms to do program
  optimisations.
\end{abstract}

@@

To preserve the generality of hylomorphisms, we encode data structures using
polynomial functors. To ensure that the extracted code is idiomatic and does not
contain unsafe casts, we avoid using indexed types in the definition of the
recursion schemes, and make sure that the definitions can be inlined by Rocq's
extraction mechanism. To strengthen the results of this paper, we avoid all
axioms.

We list the contributions of this paper. First, in
Section~\ref{sec:category-theory}, we provide the theoretical background on
recursion schemes and hylomorphisms. We then provide a Rocq formalisation of
\emph{hylomorphisms} (Section~\ref{sec:recursion-schemes}) that (1) is
\emph{fully axiom-free}; (2) allows the extraction of idiomatic and efficient
code; and (3) can use regular Rocq equalities to do program calculation, derive
correct implementations, and apply optimisations. In
Section~\ref{sec:extraction}, we apply this framework to formalise practical
examples of divide-and-conquer, dynamic programming, and mutual recursion
algorithms. Additionally, we verify the short-cut fusion optimisation and
present the extracted optimised code.

@@

Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow $C \to
A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if it
satisfies
\begin{equation}
  \label{eq:hylo}
  h = a \circ F(h) \circ c
\end{equation}
A solution to this equation does not exist for an arbitrary algebra and
coalgebra pair and, in fact, a recursive function definition like this is not
accepted by Rocq.

@@

An example of a recursive coalgebra is the \coq{partition} function used in
quicksort, which has the type $\text{List } A \to B(\text{List } A)$, where the
functor $B$ is defined as $BX = 1 + X \times A \times X$. This function deconstructs
a list by selecting a pivot element of type $A$ and splitting the remaining
elements into two sublists.

Notice that the initial algebra for lists corresponds to the functor
$F X = 1 + A \times X$ and its associated algebra map (or its inverse, viewed as
a coalgebra), $\InOp$ has the type $\text{List } A \to F(\text{List
} A)$. Importantly, \coq{partition} is not this map, which means that it does
not arise from the standard initial algebra structure, and thus catamorphisms
here cannot be used to define this recursive function.

@@

For example, in the category \textbf{Set}, consider the polynomial functor
corresponding to
\[
F(X) = A + B \times X + C \times X^2.
\]
We can model this functor by using a \emph{container}~\cite{AbbottAG05}. This is a set $S$
of \emph{shapes}, for example, $S = A + B + C$ for the example above, together with a function $P(s)$ denoting the exponent of $X$ for each shape.
Then a container extension is defined as the sum of all maps
$P(s) \to X$
\[
F(X) = \sum_{s \in S} (P(s) \to X)
\]
The functorial action
of $F$ is given by post-composition, mapping an element \( (s, g) \) in
\( F(A) \) to \( (s, f \vcomp g ) \) in \( F(B) \), for $f : A \to B$.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
%Recursion schemes provide an abstract way to consume and generate data.
We now consider the formalisation of hylomorphisms in Rocq. We first formalise
container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}). In
Section~\ref{sec:coq-coalg} we formalise coalgebras and put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}
A direct encoding of containers as presented in
Section~\ref{sec:category-theory} requires the use of the functional
extensionality axiom, and axiom K to deal with equalities of objects of type
$P(s) \to X$. Axiom K states that a predicate on equality proofs holds for all
such proofs if it holds for reflexivity. This would be needed to reason about
the equality of any two $f : P(s_1) \to X$ and $g : P(s_2) \to X$, when $s_1 =
s_2$. In Rocq, if we know $(s_1 , f) = (s_2 , g)$, we cannot extract a proof that
$f = g$ unless we assume axiom K, or the equality on the type of $s_1$ and $s_2$
is decidable.  Furthermore, upon extraction Rocq will need to insert unsafe casts
for
anything of type $P(s)$, because OCaml requires the input value to have a type that is identical to the function parameter, and
in the general case this cannot be done in OCaml.
To avoid these problems, we use \emph{setoids}, and
encode families of positions using \emph{decidable validity predicates}.

@@

The \coq{LFix F} setoid equality is the least fixed point of the \coq{App F}
setoid equality, i.e. if \coq{LFixR F} represents the setoid equality of
\coq{LFix F}, and we have two \coq{x y : LFix F}, then \coq{LFixR F x y} iff
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p'
   -> LFixR F (cont x p) (cont y p'))
\end{coqcode}
We define smart constructors for the isomorphism of least fixed
points as proper morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}
Catamorphisms are constructed so that they structurally deconstruct the
datatype, call themselves recursively, and then compose the result using an
\coq{F}-algebra.
\begin{coqcode}
Definition cata_f (a : Alg F A) : LFix F -> A
  := fix f (x : LFix) := match x with
        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
        end.
\end{coqcode}
It is easy to prove that catamorphisms respect setoid equalities.  We
define \coq{cata} as a setoid morphism between the setoid of \coq{F}-algebras and the
setoid of functions from \coq{LFix F} to \coq{A}:
\begin{coqcode}
cata : forall `{setoid A}, Alg F A ~> (LFix F ~> A)
\end{coqcode}
Finally, we prove that catamorphisms satisfy their universal property (see
Section~\ref{sec:category-theory}):
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

@@

As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra $a$ and
$F$-coalgebra $c$, $f$ is a hylomorphism if it satisfies
\[
  f = a \circ F\; f \circ c.
\]
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, a recursive function definition
like this cannot be directly accepted by Rocq.

@@

% Note that such proofs will be erased when extracting code to OCaml, and this is
% standard practice when building general recursive functions in
% Rocq\footnote{\dcas{This technique had a name in the community, but I cannot
% remember. Something predicates}}. The core definitions are the following:
% \begin{coqcode}
% Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

% \end{coqcode}
% Here, \coq{RecF_inv} provides the structurally smaller element for
% the Rocq recursive definition. We then couple coalgebras with a proof that they
% are recursive, and add a coercion from recursive coalgebras to coalgebras.
% \begin{coqcode}
% Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
% \end{coqcode}
% In some cases, proving that coalgebras can only be applied finitely many
% times is straightforward (e.g.\ they are recursive on some inductive type in
% Rocq). However, in many cases, proving that a coalgebra is recursive directly
% can be tedious and cumbersome. To ease development, we prove that if a
% given coalgebra respects a \emph{well-founded relation}, then it is
% necessarily recursive.
% \begin{coqcode}
% Definition respects_relation `{setoid A} (c : Coalg F A)
%       {B} (m : A -> B) (R : B -> B -> Prop)
%   := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).

@@

\section{Extraction}
\label{sec:extraction}
In this section we show how programs and their optimisations can be encoded in
our framework and how can they be extracted to idiomatic functional code.
Although our examples use OCaml as target language, extraction works equally
well to other target languages (e.g. Haskell or Scheme). In particular, we show
how to formalise sorting algorithms by looking at quicksort
(Section~\ref{sec:sorting}), we show how to formalise dynamic programming
algorithms by looking at knapsack (Section~\ref{sec:knap}), and, finally, we
show an example of the shortcut deforestation optimisation
(Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Rocq's generalised rewriting and the laws of morphisms to
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Rocq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

The proof is by induction on the fact that \coq{qsplit} is recursive. This means
that \coq{qsplit} terminates on \coq{l}.  In practice this is done by using the
tactic \coq{induction (recP qsplit l)}. Then we can unfold \coq{hylo merge qsplit} to
\coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
uniqueness (Section~\ref{sec:coq-rec-coalgebras}). The rest of the proof is standard.
Overall, our proof is shorter than alternative implementations, and of
comparable complexity%
%% Commenting out footnote to save space, since we mention the Rocq QSort in related work already...
\footnote{The QSort definition and correctness proof in the
equations
package is under 200LOC --
\href{https://github.com/mattam82/Rocq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Rocq-Equations/blob/8.20/examples/quicksort.v},
and other online proofs of correctness are of similar complexity
\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
% and can be found in the
% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

@@

% Mergesort is a hylomorphism with the recursive structure defined by the
% container \coq{(TreeF (list int) unit)}.
% The computationally expensive part of the Mergesort implementation is the
% algebra, since this is where most of the comparisons happen. The recursive
% coalgebra for Mergesort is defined as the following Rocq functions:
% \begin{coqcode}
% Fixpoint splitL (x : list int) (accL accR : list int) :=
%   match x with
%   | nil => (accL, accR)
%   | cons x xs => splitL xs accR (cons x accL)
%   end.

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism, and
\coq{times_two} multiplies every element of the list by two. We can use Rocq's
generalised rewriting~\cite{Sozeau09} and \coq{hylo_fusion_l} to fuse \coq{times_two} into
\coq{hylo merge qsplit} which gives the program \coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by applying
\coq{times_two} to the shapes thus multiplying every pivot by two.

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Rocq rewrite tactics.

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
We now showcase the shortcut deforestation optimisation on lists.  This is used
to remove the creation of intermediate data structures by fusing multiple
recursive traversals into one.  For example, we can use it to show that the following
code from Takano and
Meijer~\cite{TakanoM95}, which uses three different hylomorphisms, can be fused into one:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here \coq{sf1} is the composition of \coq{length},
\coq{Lmap f} and \coq{append ys} which are, in particular,
catamorphisms.
To prove our goal we use the so called \emph{acid rain} theorem~\cite{TakanoM95}
which states that given a parametric function of the following type:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
we can conclude, by parametricity, that the following equation holds:
\begin{coqcode}
  hylo a l_out \o hylo (s l_in) c =e hylo (s a) c
\end{coqcode}
Unfortunately, this is not provable in Rocq without adding the parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}, but we can prove it for specific
functors, e.g. the functor of lists.

We prove a specialised version of the acid rain theorem that fuses
hylomorphisms defined in terms of the following function \coq{tau} that maps
list algebras to list algebras.  This essentially is a function on lists which
uses the algebra for both the recursive step and for the base case where it is
used in the hylomorphism to continue the recursion:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
\end{coqcode}
Our acid rain theorem is stated as follows:
\begin{coqcode}
  hylo a l_out \o hylo (tau l l_in) c =e hylo (tau l a) c
\end{coqcode}
Now we use \coq{tau} to define the \coq{append} function as follows:
\begin{coqcode}
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Rocq lists to the
\coq{ListF} container. The acid rain theorem allows us to fuse this definition
of append with the maps \coq{Lmap} and \coq{length} which is the optimisation we
wanted in the first place. The following is the extracted OCaml code from the
optimised program:
\begin{ocamlcode}
let rec sf1 f ys =
  function | [] -> let rec f0 = function
                   | [] -> (Uint63.of_int (0))
                   | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                   in f0 ys
           | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}
Notice how the \coq{Lmap} is optimised away and the function \coq{length} is
applied to both the arguments of the append function.

@@

Our final container extension is the following:
\coq{Sum (Const nat) (Prod Id Id)}.
The shape of this functor is:
\coq{nat + unit * unit}. The positions of this functor
are \mintinline[breaklines]{Rocq}{False + (unit + unit)}, which is isomorphic
to \coq{unit + unit}, which will only be valid in the RHS.
